---
title: "Whitebark Pine Restoration Synthesis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Libraries
packageList <- c(
  "data.table", # data manipulation
  "sf", # spatial data
  "dplyr", # data manipulation
  "leaflet", # interactive maps
  "viridis", # creates colours for leaflet::colorFactor
  "fontawesome", # icons for leaflet maps
  "leaflegend", # legend for awesome icons
  "htmltools" # html code for leaflet maps
)
# Check you have them in your library
new.packages <- packageList[!(packageList %in% installed.packages()[,"Package"])]
# load them
if(length(new.packages)) install.packages(new.packages)
lapply(packageList, require, character.only = TRUE)
rm(packageList, new.packages)


# Load data
# Collection trees
alltrees <- fread("./Inputs/Seed Collection - All years.csv") # check this is the most recent version if it has been a while since last used
alltrees[Screening == "", Screening := "N"]

# Planted areas - spatial
plant.sf <- st_read(dsn = "./Inputs/AllPlantingLocations.gpkg")
plant.sf <- st_zm(plant.sf, drop=TRUE) # remove the z dimension or it won't plot
# Planted areas - data
plant <- fread("./Inputs/Planting_AllYears.csv") # IF updated Oct 26, 2023

# ARU sites
ARU <- fread("./Inputs/WBP-ARU_data_AllYears.csv")
ARU.sf <- st_as_sf(ARU, coords = c("Longitude", "Latitude"),
                   crs = "+proj=longlat")

# Camera sites
cam <- fread("./Inputs/CamSetupData2023.csv")
cam.sf <- st_as_sf(cam, coords = c("Longitude", "Latitude"),
                   crs = "+proj=longlat")

# WBP range
range.sf <- st_read(dsn = "./Inputs/WBP_Range_clason.gpkg")
range.sf <- st_transform(range.sf, crs = "+proj=longlat")

```

This document presents the work the Bulkley Valley Research Centre (BVRC) has completed for the Whitebark Pine Program since 2011. HCTF has provided funding from \_\_ to \_\_. This funding as allowed us to collect cones from rust resistant whitebark pine trees .. BLAH BLAH BLAH.

## Seed collection

In 2011, 2013, 2018, and 2022 we collected cones from trees that were identified as healthy in areas with high blister rust mortality. The cones were shucked to access the seeds, which we then send to BC Tree Centre for storage, Kalamalka Research Station for rust screening, and nurseries for future tree planting.

```{r cars, echo=FALSE}
# COLLECTION TREES

# Number of collected trees by year
filtcolltrees <- unique(alltrees$Tree_ID)
# remove the seedlots and bulked seeds from the unique collection trees
filtcolltrees <- filtcolltrees[!grepl("seedlot|bulk", filtcolltrees, ignore.case=TRUE)]
filtalltrees <- alltrees[Tree_ID %in% filtcolltrees]
colltrees <- filtalltrees[, .(NumCollTrees = uniqueN(Tree_ID)), by=YearColl]

# kg of seed collected
alltrees[Total_Seeds_g == "", Total_Seeds_g:= NA]
collSummary <- alltrees[!is.na(Total_Seeds_g), .(Seeds_g = sum(Total_Seeds_g)), by=YearColl]
collSummary[, Seeds_kg := Seeds_g*0.001] # convert g to kg

collSummary <- merge(collSummary, colltrees)
collSummary[, Seeds_g:=NULL]
collSummary

# Number of trees in the screening program
screening <- alltrees[Screening == "Y", .(Screening = uniqueN(Tree_ID))]

# Number of trees identified as elite trees
elite <- alltrees[Elite_tree == "Y", .(Elite = uniqueN(Tree_ID))]
# Number of plus trees *not sure which trees are not plus trees*
```

We have submitted `r screening` seedlings to the rust screening program. Of these screened seedlings, `r elite` have been identified as elite trees.

## Restoration

In 2011, 2012, 2014, 2017, 2021, and 2023 we planted whitebark pine seedlings in areas that had recently burned, various elevation bands, and ... Within these restoration plantations we established monitoring trees to be measured and assess the health 1, 5, 10, 15 years post planting. ** get more info on 10 year results from Sybille **

```{r pressure, echo=FALSE}
# Area planted
plant[, Area_ha := Area_m2*0.0001]

# Tree monitoring *Sybille's results*
```

## Future Directions

We would like to better understand the relationship between wildlife and whitebark pine. We know that Clark's nutcrackers are essential for planting seeds and therefore success of whitebark pine establishment, but little is known about the Clark's nutcracker at the northern limits of the whitebark pine range. Studies from the Yellowstone region have shown a strong relationship between grizzly bear diet and whitebark pine seeds, but similar to the Clark's nutcracker, there has been no research on these relationships at the northern limits.

We have started a pilot project using ARUs to observe when the nutcrackers are in a whitebark pine area and how long they stay, and cameras to observe what wildlife is using the whitebark pine ecosystems. We have identified areas which require more funding to establish proper research.

### Restoration Synthesis Map

This map presents all of the whitebark pine restoration activities that the BVRC has carried out. This includes cone collection, rust screening, and planting seedlings, as well as current habitat use projects consisting of ARUs for Clark's nutcracker detection and cameras for wildlife detection.

```{r echo=FALSE, out.width="800px", out.height="600px"}
# collection trees: column for icon/colour
alltrees[, ScreenResults := ifelse(Elite_tree == "Y", "Elite",
                                   ifelse(Screening == "Y", "Screened", "Not screened"))]
#remove rows with missing lat long - they are the bulk lots
temp <- alltrees[!is.na(Longitude) & Longitude != ""]
# convert dt to sf
alltrees.sf <- st_as_sf(temp, coords = c("Longitude", "Latitude"), 
                        crs = "+proj=longlat")
alltrees.sf$YearColl <- as.factor(alltrees.sf$YearColl)

# planted: 
plant.sf$Area <- as.numeric(plant.sf$Area)
plant.sf$Area_ha <- round(plant.sf$Area*0.0001, digits = 1)

#colours based on year planted
plant.sf$Year <- as.factor(plant.sf$YrPlntd)
colour <- colorFactor(
  palette = c("purple", "blue", "red", "pink", "orange", "green"),
  domain = c("2011", "2012", "2014", "2017", "2021", "2023")
)

# create colours and icons based on screened and elite trees
# list of icons
treeIcons <- awesomeIconList(
  "Elite" = makeAwesomeIcon(
    icon = "star",
    library = "fa",
    markerColor = "purple",
    iconColor = "black"
  ),
  "Screened" = makeAwesomeIcon(
    icon = NA,
    library = "fa",
    markerColor = "purple"
  ),
  "Not screened" = makeAwesomeIcon(
    icon = NA,
    library = "fa",
    markerColor = "blue"
  )
)

ARUicons <- makeAwesomeIcon(
  text = fa("crow"), # for some reason have to write it like this for crow to show up
  library = "fa",
  markerColor = "orange"
)

cameraIcons <- makeAwesomeIcon(
  icon = "camera",
  library = "fa",
  markerColor = "orange"
)

# Create habitat use icons for legend
habLegendIcons <- awesomeIconList(
  "ARU sites" = makeAwesomeIcon(
    text = fa("crow"),
    icon = NA,
    library = "fa",
    markerColor = "orange",
    iconColor = "black"
  ),
  "Camera sites" = makeAwesomeIcon(
    icon = "camera",
    library = "fa",
    markerColor = "orange",
    iconColor = "black"
  )
)


# Map
p <- leaflet() %>%
  addProviderTiles(providers$Esri.WorldImagery) %>% # ESRI satelite imagery
  # 2011 collection trees
  addAwesomeMarkers(data=alltrees.sf[alltrees.sf$YearColl == 2011,],
                    icon = ~treeIcons[ScreenResults],
                    popup = ~paste(Tree_ID,
                                   "<br>Year collected:", YearColl,
                                   "<br>Location:", Location),
                    group = "2011 Seed Collection") %>%
  # 2013 collection trees
  addAwesomeMarkers(data=alltrees.sf[alltrees.sf$YearColl == 2013,],
                    icon = ~treeIcons[ScreenResults],
                    popup = ~paste(Tree_ID,
                                   "<br>Year collected:", YearColl,
                                   "<br>Location:", Location),
                    group = "2013 Seed Collection") %>%
  # 2018 collection trees
  addAwesomeMarkers(data=alltrees.sf[alltrees.sf$YearColl == 2018,],
                    icon = ~treeIcons[ScreenResults],
                    popup = ~paste(Tree_ID,
                                   "<br>Year collected:", YearColl,
                                   "<br>Location:", Location),
                    group = "2018 Seed Collection") %>%
  # 2022 collection trees
  addAwesomeMarkers(data = alltrees.sf[alltrees.sf$YearColl == "2022",],
                    icon = ~treeIcons[ScreenResults],
                    popup = ~paste(Tree_ID,
                                   "<br>Year collected:", YearColl,
                                   "<br>Location:", Location),
                    group = "2022 Seed Collection") %>%
  # ARU sites
  addAwesomeMarkers(data = ARU.sf,
                    icon = ARUicons,
                    popup = ~paste("<br>Location:", Location,
                                   "<br>Deployment year:", Year),
                    group = "ARU sites") %>%
  # Camera sites
  addAwesomeMarkers(data = cam.sf,
                    icon = cameraIcons,
                    popup = ~paste("<br>Location:", Location,
                                   "<br>Deployment year:", Year),
                    group = "Camera sites") %>%
  # # add layers on/off control
  addLayersControl(overlayGroups = c("2011 Seed Collection",
                                     "2013 Seed Collection",
                                     "2018 Seed Collection",
                                     "2022 Seed Collection",
                                     "Planting Locations",
                                     "ARU sites",
                                     "Camera sites",
                                     "WBP range"),
                   position = "topright") %>%
  # Collection tree legend
  addLegendAwesomeIcon(
    iconSet = treeIcons,
    title = "Collection Trees",
    position = "topright",
    orientation = "vertical") %>%
  # Habitat use legend
  addLegendAwesomeIcon(
    iconSet = habLegendIcons,
    title = "Habitat Use",
    position = "topright",
    orientation = "vertical") %>%
  #WBP range
  addPolygons(data = range.sf,
              color = "green",
              fillOpacity = 0.25,
              group = "WBP range") %>%
  # Restoration plant areas
  addPolygons(data = plant.sf,
              color = ~colour(Year),
              popup = ~paste(Name,
                             "<br>Year planted:", Year,
                             "<br> Area (ha):", Area_ha),
              fillOpacity = 1,
              highlightOptions = highlightOptions(color = "white",
                                                  weight = 2,
                                                  bringToFront = TRUE),
              group = "Planting Locations") %>%
  # Planting colour legend
  addLegend("topright", pal = colour, values = plant.sf$Year,
            title = "Year Planted",
            opacity = 1)

p
```
